### 2022-09-06 main/chapter_searches.tex
pvalue ratio versus likelihood ratio

Posterior probabilities are, however, useful when digesting data to make
inferences about the world.
Reporting likelihood functions and evidence values allows readers to assign
priors as they wish.
For model comparison, for example between background $h=B$ and signal $h=S$
models, the analysis is conveniently linearized by expressing it in $\log$
ratio form:
\begin{equation}
\hphantom{\quad \quad \mid\mid \mathrm{context}}
\phi(d) =
\log\frac{\prob{S}{d}}{\prob{B}{d}}
=
\log\frac{\prob{d}{S}}{\prob{d}{B}}
+
\log\frac{\probc{S}}{\probc{B}}
\quad \quad \mid\mid \mathrm{context}
,
\end{equation}
where we use the double bar, $\mid\mid$, to refer to any contextual information
a reader might wish to include.
When the evidence ratio is independent of that context, as it usually is,
it factors out from the prior (odds) ratio.
In this sense, likelihood (or evidence) ratios are scale factors that update
any odds prior ratio by the same multiplication, and that on this $\log$ scale
that means just adding a constant.

From such $\log$ ratios one individual probabilities are recovered through the
logistic function:
\begin{equation}
\label{eqn:searches_logistic}
\prob{S}{d} = \frac{1}{1 + e^{-\phi(d)}}
.
\end{equation}
Physically, this is analogous to the Boltzmann distribution where the energy
of each model is its negative $\log$ probability, with unit inverse
temperature~\cite{
pmlr-v2-ranzato07a,
skilling2017david
}.
With multiple models in play, it is also called the `softmax'
function~\cite{MurphyKevinP.2012Mlap}.

@@

Some literature names this Fisher/Neyman-Pearson assemblage
``classical'' statistics~\cite{lehmann2011fisher, Feldman:1997qc}, but that is
simply inaccurate for a 20th century creations that innovated on 18th century
ideas~\cite{bayes1763lii, laplace1774stigler}.
Jaynes instead names them ``orthodox''~\cite{jaynes2003probability},
which may be accurate in some contexts, but alternative
(heretical?) ideas have taken hold in research at the LHC ---
the orthodoxy has moved on.
We therefore stick with `frequentist'.

@@

% test statistics
A statistic is a function of data, and a test statistic is just a statistic
that is used in statistical tests.
For \atlas\ collision data, event variables are statistics; when working with
data, one always uses statistics and is free to design them for their benefits
in context.
Just as we define event variables (like $\pt$) to remove nuisance
dependencies (on $\phi$ rotations or longitudinal boosts),
useful test statistics can also eliminate unwanted parameters.
The quintessential example is $t$-statistic
\begin{equation}
t(d) = \frac{\bar{\mu}(d) - \mu}{\bar{\sigma}(d)}\sqrt{n}
,
\end{equation}
where $\mu$ is the known data mean, and $\bar{\mu}(d)$ and $\bar{\sigma}(d)$
are standard estimates of the mean and standard deviation of the data sample
respectively~\cite{student1908, fisher1925t}.
By dividing out a (scale equivariant) standard deviation, $t$ is independent
of common scale factors on the data, and that scale invariance has helped to
drive the study and common use of this $t$ statistic~\cite{lehmann2011fisher}.

One can usefully reduce observed data to just the test statistic
$\tilde{t} = t(d)$ and compare likelihoods to compare how well different
hypotheses predict the observed $\tilde{t}$.
For reasons I cannot explain, Fisher advocates a different use ---
rather than considering likelihoods and likelihood ratios, he recommends
that researchers consider the ``probable error''~\cite{fisher1921probable},
which is a cumulative distribution function in the negated test statistic.
In modern jargon, a `probable error' at the observed test statistic
$\tilde{t}$ and an hypothesis $h$ is a \textbf{$p$-value}
\begin{align}
P(h)
&= \prob{t \geq \tilde{t}}{h} \\
&= L(h) + \prob{t > \tilde{t}}{h}
\label{eqn:searches_pvalue_likelihood}
;
\end{align}
the $p$-value $P(h)$ is the prior probability of data
\emph{greater or equal to} the observed data when ordered by the value of the
test statistic.
Equation~\ref{eqn:searches_pvalue_likelihood} emphasizes that it is just the
likelihood plus some fluff from a tail of larger test statistics.

\TODO{quote EX.II from ~\cite{fisher1921probable}
}

>>>>

Posterior probabilities are, however, useful when digesting data to make
inferences about the world.
Reporting likelihood functions and evidence values allows readers to assign
priors as they wish.
For model comparison, for example between background $h=B$ and signal $h=S$
models, the analysis is conveniently linearized by expressing it in $\log$
ratio form:
\TODO{different choice for phi to avoid angle?}
\begin{equation}
\hphantom{\quad \quad \mid\mid \mathrm{context}}
\phi(d) =
\log\frac{\prob{S}{d}}{\prob{B}{d}}
=
\log\frac{\prob{d}{S}}{\prob{d}{B}}
+
\log\frac{\probc{S}}{\probc{B}}
\quad \quad \mid\mid \mathrm{context}
,
\end{equation}
where we use the double bar, $\mid\mid$, to refer to any contextual information
a reader might wish to include.
When the evidence ratio is independent of that context, as it usually is,
it factors out from the prior (odds) ratio.
In this sense, likelihood (or evidence) ratios are scale factors that update
any odds prior ratio by the same multiplication, and that on this $\log$ scale
that means just adding a constant.

As data are reported more precisely, likelihoods will often tend towards zero
--- in the continuum limit, the probability of any data value vanishes.
Small likelihoods are themselves not significant unless other likelihoods
are larger --- unless other hypotheses predicted \emph{better}.
And these ($\log$) ratios can remain well defined in continuum limits even as
both likelihoods tend towards zero~\cite{billingsley2008probability}.

From such $\log$ ratios, a little rearrangement recovers the individual
probabilities through the logistic function:
\begin{equation}
\label{eqn:searches_logistic}
\prob{S}{d} = \frac{1}{1 + e^{-\phi(d)}}
.
\end{equation}
Physically, this is analogous to the Boltzmann distribution where the energy
of each model is its negative $\log$ probability, with unit inverse
temperature~\cite{
pmlr-v2-ranzato07a,
skilling2017david
}.
With multiple models in play, it is also called the `softmax'
function~\cite{MurphyKevinP.2012Mlap}.

@@

Some literature names this Fisher/Neyman-Pearson assemblage
``classical'' statistics~\cite{lehmann2011fisher, Feldman:1997qc}, but that is
inaccurate for 20th century creations that innovated on 18th century
ideas~\cite{bayes1763lii, laplace1774stigler}.
Jaynes instead names them ``orthodox''~\cite{jaynes2003probability},
which may have been accurate in his environment, but different
(heretical?) ideas have taken hold in research at the LHC ---
the orthodoxy has moved on.
We therefore stick with `frequentist'.

@@

% test statistics
A statistic is a function of data, and a test statistic is just a statistic
that is used in statistical tests.
For \atlas\ collision data, event variables are statistics; when working with
data, one always uses statistics and is free to design them for their benefits
in context.
Just as we define event variables (like $\pt$) to remove nuisance
dependencies (on $\phi$ rotations or longitudinal boosts),
useful test statistics can also eliminate unwanted parameters.
The quintessential example is the $t$-statistic
\begin{equation}
t(d) = \frac{\bar{\mu}(d) - \mu}{\bar{\sigma}(d)}\sqrt{n}
,
\end{equation}
which shifts the sample mean $\bar{\mu}(d)$ by a known true mean $\mu$, and
scales that difference down by the same standard deviation
$\bar{\sigma}(d)$~\cite{student1908, fisher1925t}.
By dividing out a scale equivariant estimate of the standard deviation, $t$ is
independent of common scale factors on the data, and that scale invariance
has helped to drive the study and common use of this
$t$-statistic~\cite{lehmann2011fisher}.

One can usefully reduce observed data to just the test statistic
$\tilde{t} = t(d)$, and compare likelihoods on that observed $\tilde{t}$
to compare how well different hypotheses predict it.
For reasons I cannot explain, Fisher advocates a different use ---
rather than considering likelihoods and likelihood ratios, he recommends
that researchers consider the ``probable error''~\cite{fisher1921probable},
which is a cumulative distribution function in the negated test statistic.
In modern jargon, a `probable error' at the observed test statistic
$\tilde{t}$ and an hypothesis $h$ is a \textbf{$p$-value}
\begin{align}
P(h)
&= \prob{t \geq \tilde{t}}{h}
\label{eqn:searches_pvalue}
\\
&= L(h) + \prob{t > \tilde{t}}{h}
\label{eqn:searches_pvalue_likelihood}
;
\end{align}
the $p$-value $P(h)$ is the prior probability of data
\emph{greater or equal to} the observed data when ordered by the value of the
test statistic.
Equation~\ref{eqn:searches_pvalue_likelihood} points out that the $p$-value is
just is the likelihood plus some fluff from a tail of larger test statistics.

Fisher illustrates the use of $p$-values in the second example
of~\cite{fisher1921probable}, which models some data with a parameter $r$ and
compares two hypotheses
$r=0.3$ and
$r=0.95$, finding $p$-values
$P(r=0.3)=0.142$ and
$P(r=0.95)=0.00014$.
``Thus'', Fisher concludes, ``the value $0.95$\ldots
is roughly $1000$ times less likely than $0.30$''
(``when measured on the $r$ scale'').
By writing `likely', Fisher is carefully not referencing the probability
of parameters, but their likelihoods in our definition.
But this factor of $1000$ is a $p$-value ratio, not a likelihood ratio,
so how can this make sense?
Plausibly, it is a data reduction; comparing
Equations~\ref{eqn:searches_pvalue} and~\ref{eqn:searches_likelihood}
shows that the $p$-value is a likelihood (only) if the result is reduced from
$A: t =    \tilde{t}$ to
$B: t \geq \tilde{t}$.
And since this reduction is technically true --- $A$ implies $B$ --- the
likelihood ratio on $B$ is logically valid and not \emph{necessarily}
misleading.

But $B: t \geq \tilde{t}$ is a logical statement (true for all test statistics
greater than or equal to the observed value) and not a randomly occurring
event, so its assignment frequentist probability is deeply dubious.
Much more importantly, the reduction to a one-sided inequality has no tangible
benefit; it reduces the information content of the reported data, but does not
reduce the costs of reporting or interpreting those data.
Both $A$ and $B$ are printed in the same number of characters, and $p$-values
can be painfully expensive to evaluate.
Furthermore, it can actively harm the result by diluting the likelihood ratio,
since if
\begin{equation}
\phi = \log \frac{L(h_1)}{L(h_2)}
\to
\log \frac{
L(h_1) + \prob{t > \tilde{t}}{h_1}
}
{
L(h_2) + \prob{t > \tilde{t}}{h_2}
}
\end{equation}
then large tail probabilities will only squash the $\log$ ratio toward zero.
